---
layout: post
title:  "Data Driven Decision-making"
categories: [data, academic life, decision, priorities, ranting]
---

```{r echo=FALSE, message=FALSE}
knitr::opts_knit$set(base.dir = "/home/swolf/website/", base.url = "/")
knitr::opts_chunk$set(fig.path = "figure/2022-02-25-dataDrivenDecisions/", 
                      fig.width=10,fig.height=6)

library(ggplot2)
library(viridis)
```

I have been thinking lately what it means to engage in data-driven decision making from an institutional perspective.  One of the things that I have heard university leadership state over the years is that we make decisions based on data.  I happen to be a [Faculty Senator](https://facultysenate.ecu.edu) at the time of this writing, and the more that I interact with that body, the more that I question whether this institution indeed meaningfully engages with the data that we generate.

## Data analysis as counting.

In our [February meeting](https://www.ecu.edu/cs-acad/fsonline/customcf/fsagenda/2022/fsa222.pdf), the faculty senate had the pleasure of listening to a presentation about grades over the past year. This report is typical of the standard that we see for "data analysis" in these reports, and I'm picking this report to criticize for reasons that include it being fresh in my mind and that I have some interest and expertise in student grades given my research field.  I don't think badly of the presenter, nor do I think that the presentation was badly done.  I just don't think that this report was terribly valuable to help the Faculty Senate engage in data driven decision making as an informed member of our shared governance structure.  Let's talk about what I mean by that.

This report is typical of what I could call "data analysis as counting."  We were given a report on university grades updated over the past year.  We saw a number of plots that looked like this one showing grade distributions for the entire university.

```{r echo=FALSE}
gradeDistData = read.csv('ecuGradeData/ecuGradeData.csv')
gradeDistData$grade = factor(gradeDistData$grade)

ggplot(gradeDistData, aes(fill = grade, y = gradeNum, x = acadYear)) +
    geom_bar(position=position_fill(reverse = TRUE), stat='identity') +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_viridis(discrete = T) + ylab('Percentage of grades') +
    xlab('Academic Year')
```

This was then repeated for one of the colleges, one of the departments, and one of the courses.  And they all, mostly, looked the same, at least to my eye.  (The explosion of `P` grades in 2019-2020 and 2020-2021 is explained by the fact that we allowed students to be graded on a PASS/FAIL basis due to the COVID pandemic and the subsequent abrupt transitions to online instruction.)




